---
title: "Reproducible Research: Peer Assessment 1"
author: "Thomas Fischer"
date: "March 7, 2018"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(base.dir = "figure")

options(scipen=999)
```

## Loading and preprocessing the data
**1. Load the data (i.e. read.csv())**
```{r load_data, cache=TRUE}
if(!file.exists('activity.csv'))
    unzip(zipfile = 'activity.zip')
activity <- read.csv('activity.csv',
                     header = TRUE,
                     stringsAsFactors = FALSE)
```
**2. Process/transform the data (if necessary) into a format suitable for your analysis.**  
Converting character date variable to object of class "Date"
```{r convertToDate}
activity$date <- as.Date(activity$date)
```
and mergeing date and interval into a "Date-Time" variable
```{r convertToDateTime, message=FALSE, warning=FALSE, results='hide'}
# activity$time <- sub('^([0-9]{1})$', '0:0\\1', activity$interval)
# activity$time <- sub('^([0-9]{2})$', '0:\\1', activity$time)
# activity$time <- sub('^([0-9]{1})([0-9]{2})$', '\\1:\\2', activity$time)
# activity$time <- sub('^([0-9]{2})([0-9]{2})$', '\\1:\\2', activity$time)
# more elegant
activity$Time <- sprintf("%04d", activity$interval)
activity$Time <- sub('^([0-9]{2})([0-9]{2})$', '\\1:\\2', activity$Time)

require(lubridate)
activity$DateTime <- ymd_hm(paste(activity$date, activity$Time))
activity$Time <- ymd_hm(paste('1900-01-01', activity$Time))
activity$day <- weekdays(activity$date)
```
Now let's have a first look at our data:
```{r first_look}
str(activity)
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.  
**1. Calculate the total number of steps taken per day**
```{r calculate_StepsPerDay}
stepsbyday<- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
```
**2. Make a histogram of the total number of steps taken each day**
```{r plot_histogramm, fig.height = 5, fig.width= 8}
hist(stepsbyday,
     breaks = 26,
     col = 'blue',
     main = 'Histogram of Steps per Day',
     xlab = 'Steps per Day',
     ylab = 'Number of Days',
     xlim = c(0,22000))
```

**3. Calculate and report the mean and median of the total number of steps taken per day**
```{r calc_centers}
mean_stepsbyday <- round(mean(stepsbyday),0)
median_stepsbyday <- median(stepsbyday)
```
Steps per Day|Value
-----------|-----
Mean|`r round(mean_stepsbyday,2)`
Median|`r median_stepsbyday`

## What is the average daily activity pattern?
**1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**  
First calculate the average number of steps per 5-minute interval:
```{r calc_mean_per_interval, message=FALSE, warning=FALSE, results='hide'}
require(dplyr)
MeanStepsTime <-
    activity %>%
    group_by(Time) %>%
    summarize(Mean = round(mean(steps, na.rm = TRUE),1)) %>%
    as.data.frame()
```
and then plot it:
```{r plot_mean_per_interval, fig.height = 5, fig.width= 8}
plot(MeanStepsTime, 
     type = 'l',
     lwd = 1,
     main = 'Average Number of Steps over Time Intervals',
     xlab = 'Time Interval',
     ylab = 'Average Number of Steps')
```  

**2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r}
maxStepsInterval <-
    strftime(MeanStepsTime[which.max(MeanStepsTime$Mean),1],
             format = '%H:%M', tz = 'UTC')
maxSteps <- MeanStepsTime[which.max(MeanStepsTime$Mean),2]
```
The interval at `r maxStepsInterval` contains on average `r maxSteps` steps, which is the maximum.

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  
**1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**
```{r missing_values}
missing_observations <-
    sum(is.na(activity$steps))
total_observations <-
    length(activity$steps)
missing_percentage <-
    round(missing_observations/total_observations * 100,1)
```
There are `r missing_observations` missing observations for variable 'steps', which accounts for `r missing_percentage`% of the total of `r total_observations` observations.  

**2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**  
```{r calc_nas}
missing_by_date <- 
    activity %>%
    group_by(date) %>%
    summarize(
        observations = n(),
        num_NAs = sum(is.na(steps))) %>%
    as.data.frame()
```
Looking at the distributions we realize that there is missing data for complete days only.
```{r explore_nas}
missing_by_date[missing_by_date$num_NAs > 0,]
```
Further on there are a two strange days, namely '2012-10-02' and '2012-11-15' with only two time intervals not equal to zero. Most likely the wearable is on but wasn't used. So best strategy is to impute NA's for 0 values for this special day. Later we can impute a better value.  
```{r outlier}
activity[(activity$date == as.Date('2012-10-02') | 
              activity$date == as.Date('2012-11-15')) & activity$steps > 0,c(1,2,3,6)]
```
I would tend to just disregard the days with missing values. But as we are required in this assignment to impute missing values, the strategy will be to fill the missing values with the median for the specific weekday and time interval.  
**3.Create a new dataset that is equal to the original dataset but with the missing data filled in.**  
```{r imputed_df}
activity_imputed <-
    activity %>%
    mutate(steps = ifelse(
        (date == as.Date('2012-10-02') | date == as.Date('2012-11-15')) & 
            steps == 0,
        NA, steps))
activity_imputed <- activity_imputed %>%
    group_by(day, interval) %>%
    mutate(steps = 
               ifelse(is.na(steps), median(steps, na.rm = T), steps)) %>%
    as.data.frame()
```
**4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**  
```{r calculate_StepsPerDay_2}
stepsbyday_imputed<- tapply(activity_imputed$steps, activity_imputed$date, sum, na.rm=TRUE)
```
```{r plot_histogramm_2, fig.height = 5, fig.width= 8}
hist(stepsbyday_imputed,
     breaks = 25,
     col = 'blue',
     main = 'Histogram of Steps per Day',
     xlab = 'Steps per Day',
     ylab = 'Number of Days',
     xlim = c(0,22000))
```
```{r calc_centers_2}
mean_stepsbyday_imputed <- round(mean(stepsbyday_imputed),0)
median_stepsbyday_imputed <- as.integer(median(stepsbyday_imputed))
```
As we can see in following table the mean increased while the median remained the same. This makes sense because in our before data we faced days with 0 steps. Imputing values for these increased the mean. The median remained the same because it is stable to outliers.  

Steps per Day|Value
-----------|-----
Mean before Imputation  | `r mean_stepsbyday`
Mean after Imputation   | `r mean_stepsbyday_imputed`
Median before Imputation| `r median_stepsbyday`
Median after Imputation | `r median_stepsbyday_imputed`  
## Are there differences in activity patterns between weekdays and weekends?
**1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**  
```{r weekday_type}
normal_weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity_imputed <- 
    activity_imputed %>%
    mutate(day_type = ifelse(day %in% normal_weekdays,
                             'weekday', 'weekend'))
activity_imputed$day_type <- as.factor(activity_imputed$day_type)
```
**2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.**  
```{r calc_means}
MeanStepsTime <-
    activity_imputed %>%
    group_by(day_type, Time) %>%
    summarize(Average = round(mean(steps, na.rm = TRUE),1)) %>%
    as.data.frame()
```
Panel Plots contrasting weekdays with weekends data:
```{r plot_panel, fig.height = 10, fig.width= 8}
par(mfrow=c(2,1))
plot(MeanStepsTime[MeanStepsTime$day_type == 'weekday',2],
     MeanStepsTime[MeanStepsTime$day_type == 'weekday',3],
     type = 'l',
     main = 'Number of Steps (WeekDay)',
     xlab = 'Time Interval',
     ylab = 'Average of Steps',
     ylim = c(0,250))
plot(MeanStepsTime[MeanStepsTime$day_type == 'weekend',2],
     MeanStepsTime[MeanStepsTime$day_type == 'weekend',3],
     type = 'l',
     main = 'Number of Steps (WeekEnd)',
     xlab = 'Time Interval',
     ylab = 'Average of Steps',
     ylim = c(0,250))
```